A lottery model of density-dependent selection

### 2017-11-08 manuscript.tex
l

MacArthur considers two types subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$. Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it can be seen that one type will be excluded unless it is able to grow at higher densities than the other type in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a). MacArthur defines four ``carrying capacities'' generally labeled $K$, as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$). Assuming that the nullclines are linear, these $K$ values determine whether a region of higher-density growth exists for each type.

Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$. Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.  Thus, for MacArthur ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities. This general idea applies even if the nullclines are so curved that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

Clearly MacArthur's claim is not compatible with the constant-$N$, relative fitness description of selection, in which $n_1$ and $n_2$ change along a line defined by $n_1+n_2=N$, and population density is otherwise irrelevant (Fig.~\ref{fig:Ksel}b). In this case, all four $K$ values are equal to $N$ [not!], and the $f_1$ and $f_2$ nullclines do not even exist. Since $N$ is fixed, it is not possible to persist at higher densities than $N$, nor is it possible to suppress the density of a competitor at densities lower than $N$

>>>>

MacArthur considers two types subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$. Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it can be seen that a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. For a type to have the possibility of persisting, it must be able to grow in some region of $(n_1,n_2)$ space at higher total population density $N=n_1+n_2$ than the other type (Fig.~\ref{fig:Ksel}a).

MacArthur defines four ``carrying capacities'' as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$). Assuming that the nullclines are close to linear, these $K$ values determine whether a region of higher-density growth exists for each type. Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$. Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.  Thus, to summarize, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

Clearly MacArthur's argument is not compatible with the constant-$N$, relative fitness description of selection, in which $n_1$ and $n_2$ change along a line defined by $n_1+n_2=N$, and population density is otherwise irrelevant (Fig.~\ref{fig:Ksel}b). Since $N$ is fixed, it is not possible to persist at higher densities than $N$, nor is it possible to suppress the density of a competitor at densities lower than $N$

In this case, all four $K$ values are equal to $N$ [not!], and the $f_1$ and $f_2$ nullclines do not even exist.
